Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model

### 2014-04-19 trunk/paper_enahpc14/Introduction.tex
New version

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
to  the  investigated   performance  and  energy  evaluation  methods.
Sec.~\ref{sec:3} describes different power measurement equipments used
for  conducting  this  benchmarking  study and  presents  energy-aware
techniques  employed   to  determine  the  energy   footprint  of  the
considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
from  the Exa2Green  project and  highlights areas  where improvements
will be  necessary for the subsequent  baseline refactoring.  Finally,
we  conclude with  practical guidelines  to identifying  energy saving
opportunities  and give  directions  for further  developments of  the
modeling system in Sec.~\ref{concl}.

>>>>

This work is organized as follows: in Sec.~\ref{sec:1} we describe related work, then in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART and specify its  technical setup related to the investigated
performance and energy evaluation methods.  Sec.~\ref{sec:5} describes
the HPC supercomputers used to compute the baseline. Sec.~\ref{sec:3} describes
different  power  measurement  equipment  used  for  conducting  this
benchmarking  study and presents  energy-aware techniques  employed to
determine  the   energy  footprint  of   the  considered  application.
Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
and  highlights areas  where improvements  will be  necessary  for the
subsequent baseline refactoring.  Finally, we conclude with practical
guidelines  to identifying  energy  saving opportunities  and give  an
outlook to future research in Sec.~\ref{concl}.

### 2014-04-19 trunk/paper_enahpc14/Sec3:Measurement-framework.tex
New version

In this section,  we present two different frameworks  deployed on HPC
systems  to  measure the  power  consumption  and  performance of  the
baseline execution.

\subsection{Framework description}
\label{subsec:3.1}

\subsubsection{E3METER metering products}
Supercomputer  clusters considered  for our  experiments at  the Swiss
National Supercomputing Center of  ETH Zurich (CSCS) are equipped with
E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
high   quality   electricity  meters   released   by  Riedo   Networks
(\url{http://riedonetworks.com/}),  that  enable  to monitor  and  log
power  consumption of  the  IT infrastructure  as  well as  constantly
analyze  line  voltage,  current,  power-factor,  frequency  with  1\%
accuracy.   Using reliable  narrowband  powerline communication  (PLC)
technology,  all metering  and power  quality data  from each  IPS are
centrally collected by the E3METER Data Concentrator, via the existing
power cables thus  avoiding the need for extra  cabling.  This data is
made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
Ethernet  port.   Time  synchronization  is guaranteed  by  using  NTP
servers.   Measured data  is accessed  through the  open  source Cacti
software including  the E3METER  Cacti Plugin to  scan the  entire PLC
network and monitor  in real-time the power usage  of individual rack,
recorded in 5 minutes interval periods.

\subsubsection{University of Hamburg}
To assess the  performance and the energy efficiency  of COSMO-ART, we
employ   a    version   of   the    integrated   framework   presented
in~\cite{energy13} that works in  combination with Extrae and Paraver,
which are profiling/tracing and visualization tools, respectively.

@@

>>>>

In this section,  we present three different frameworks  deployed on HPC
systems  to  measure the  power  consumption  and  performance of  the
baseline execution.

%\subsection{Framework description}
%\label{subsec:3.1}

\subsection{E3METER metering products}
Supercomputer  clusters considered  for our  experiments at  the Swiss
National Supercomputing Center of  ETH Zurich (CSCS) are equipped with
E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
high   quality   electricity  meters   released   by  Riedo   Networks
(\url{http://riedonetworks.com/}),  that  enable  to monitor  and  log
power  consumption of  the  IT infrastructure  as  well as  constantly
analyze  line  voltage,  current,  power-factor,  frequency  with  1\%
accuracy.   Using reliable  narrowband  powerline communication  (PLC)
technology,  all metering  and power  quality data  from each  IPS are
centrally collected by the E3METER Data Concentrator, via the existing
power cables thus  avoiding the need for extra  cabling.  This data is
made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
Ethernet  port.   Time  synchronization  is guaranteed  by  using  NTP
servers.   Measured data  is accessed  through the  open  source Cacti
software including  the E3METER  Cacti Plugin to  scan the  entire PLC
network and monitor  in real-time the power usage  of individual rack,
recorded in 5 minutes interval periods.

\textbf{MALOSSI: I MOVED THIS PART HERE, SINCE IT IS RELATED TO CSCS MACHINES: IT NEEDS TO BE INTEGRATED IN THE TEXT ABOVE}
Energy  consumption of  CSCS  clusters is
assessed  by sampling  the instantaneous  peak power  during execution
which  is then  averaged  and multiplied  by  the time-to-solution  to
determine energy-to-solution.

\subsection{University of Hamburg}
To assess the  performance and the energy efficiency  of COSMO-ART, we
employ   a    version   of   the    integrated   framework   presented
in~\cite{energy13} that works in  combination with Extrae and Paraver,
which are profiling/tracing and visualization tools, respectively.

@@

\subsection{IBM BG/Q integrated power measurement system}
The IBM Blue~Gene/Q~(BG/Q) supercomputer is equipped with internal hardware sensors
that provide 8~power measures for each node board (32~node cards, 512~cores), including total power, chip power, DDR~DIMM power, and network power.
Therefore, no additional hardware tools are required to obtain power measurements.
In order to instrument the code and collect power measurements in a file we use the MonEQ library~\cite{SeanWallace2013},
which provides a simple interface to start and stop power measurement in specific regions of the code.
Energy is computed a posteriori as the product of average power and time-to-solution.

### 2014-04-19 trunk/paper_enahpc14/Sec4:Evaluation.tex
New version

In  this  section, we  present  a  performance  and energy  efficiency
evaluation  of  different  achitectures  when  running  the  COSMO-ART
baseline.  We  start by  specifying our measurement  methodology along
with the metrics used to  analyse the results on all platforms.  Then,
we detail the environment  setup gathering the considered HPC systems,
the software  environment and the runtime  configuration.  Finally, we
discuss  benchmark   results  and  power-performance   traces  of  the
application.

\subsection{Measurement methodology}
\label{subsec:4.1}
We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution}       and       \textit{energy-to-solution}.
Time-to-Solution  refers   to  the  total  wall  clock   time  of  the
application execution time. Energy-to-solution is the amount of energy
spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
assessed  by sampling  the instantaneous  peak power  during execution
which  is then  averaged  and multiplied  by  the time-to-solution  to
determine energy-to-solution.   Whenever possible, multiple production
runs of COSMO-ART were  performed to illustrate the reproducibility of
the baseline, and quantify  the significant uncertainties in the power
measurement, as dictated by the available technology.

\subsection{Environment setup}
\label{subsec:4.2}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the milestone
versions   of    COSMO-ART.    In   this    study,   a   complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
v2 ten-core processors operating at 2.2 GHz, themselves connected by a
high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
energy-to-solution benchmark,  a full rack of Monch  constituted of 52
standard  compute   nodes  (monchc[029-080])  was   considered.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
based cluster  used as Piz  Daint pre-post processing cluster.   It is
composed of 42 compute nodes  and has 2 high-speed interconnects based
on FDR:  the first is dedicated to  the MPI traffic and  the second to
the storage high speed traffic.  The 2 login nodes and the 42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
themselves  connected by  a high  speed InfiniBand  network,  based on
Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
was considered.

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both many-core  platforms was  controlled using the  modules framework
which gives  an easy and  flexible mechanism to  access to all  of the
CSCS  provided compilers,  tools  and applications.   For our  initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the  x and  y  coordinates  was defined  by  setting: $nprocx=40$  and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

\subsection{Experimental results}
\label{subsec:4.3}

>>>>

In  this section,  we  describe a  performance  and energy  efficiency
evaluation  of  different  achitectures  when  running  the  COSMO-ART
baseline.  We  start by  specifying the metrics used to analyse the results on all platforms.
Then, we detail the environment  setup gathering the considered HPC systems,
the  software  environment and  the  run  configuration.  Finally,  we
discuss benchmark  results and  power-performance traces of  the model
system.

\subsection{Measurement methodology}
\label{subsec:4.1}
We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution}       and       \textit{energy-to-solution}.
Time-to-solution  refers   to  the  total  wall  clock   time  of  the
application execution time. Energy-to-solution is the amount of energy
spent  to achieve  results.  Whenever possible, multiple production
runs of COSMO-ART were  performed to illustrate the reproducibility of
the baseline, and quantify  the significant uncertainties in the power
measurement, as dictated by the available technology.

\subsection{Sofware environment}
\label{subsec:4.2}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.

\textbf{MALOSSI: Here I suggest to create small subsections for each machine to avoid confusion especially because we need to put details of 4 different machines\ldots}

The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{IBM BG/Q}
\textbf{THIS PART NEEDS TO BE COMPLETED. IT IS JUST A PLACEHOLDER}
All tests are compiled with the IBM~XL~C/\cpp compiler, with the options: {\ttfamily\scriptsize -qarch=qp -qtune=qp -O3}, and involve double-precision arithmetic.

\subsection{Run configuration}
\label{subsec:4.3}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch, on a  1344 cores using  16 MPI  tasks per  node on
Pilatus, and on 512 cores using 32 MPI tasks per node on the IBM BG/Q.
The  calculated  region  was mapped  to  the  participating
processors using  a 2-D partitioning strategy.   \textbf{MALOSSI: I suggest to remove the following part. It adds low level details that probably are not useful for the reader at this level.}
The distribution along
the~$x$ and~$y$~coordinates  was defined  by  setting: $nprocx=40$  and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$ \textbf{(for Pilatus?)} as $nprocx$ is
usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution \textbf{Also this last sentence is quite strong: I would remove it because we might reconsider this opportunity in a future stage of optimization}.\\

\subsection{Experimental results}
\label{subsec:4.4}
